Measurement of a dwarf galaxy rotational velocity from its Lyman-alpha line emission

### 2016-05-18 paper/letter.tex
recompile
better format for the fit plot
minor edits

\begin{document}
\pagestyle{empty}
\noindent
\textbf{Þú ert jörðin}
\\
\\
JEFR$^{1}$, MCRG$^1$, JNGC$^2$, MD$^3$
\\
\\
\scriptsize
{$^1$ Bogota
\\
$^2$ Tucson
\\
$^3$ Oslo
\normalsize
\\
\\
\textbf{
Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected pristine conditions of the first galaxies.  
Until these early galaxy generations are observationally
detected, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.  
Here we report on one of such CDGs, \tol, which presents
features in its Lyman-$\alpha$ emission thad had evaded theoretical
interpretation so far. 
We show that these special features, a symmetric triple peaked
emission line, are naturally explained by gas rotation.  
We find that the Lyman-$\alpha$ emission region in \tol\ should
have a rotational velocity of $V_{r}=300$ km s$^{-1}$ and a neutral
Hydrogen column density of $\log N_{HI} / \mathrm{atoms\ cm}^{-2} =
  20.5$.   
Using archival observational information we find that
the diameter for that region should be in the range of
$110$ pc $<D<$ $340$ pc and its total dynamical mass 
should be between $2.1\times 10^{9}$M$_{\odot}$ $<M_D<$  $6.6\times
10^{9}$M$_{\odot}$. 
This dynamical mass is at least $16\pm 9$ times
larger than the neutral mass hydrogen.
We argue that a possibility to explain the excess in
dynamical mass is the presence of a super-massive black hole. }

The first generation of galaxies trace our cosmic origins. 
Their characteristiscs are the first step on the evolution of galaxies
such as the Milky Way.
In the standard Big Bang cosmology the only elements that were
created in the nucleosynthesis process were Hydrogen, Helium and
Lithium. 
Heavier elemets must have been created in stellar evolution process. 
Therefore the first generation of
galaxies should be metal free and rich in Hydrogen. 
This kind of primordial galaxies have not been detected yet. 
However, dwarf star forming galaxies with a low metallicity content
are close template to understand the early galaxy evolution process.

Almost fifty years ago \cite{PartridgePeebles} it was realized that
young galaxies could be detected through a strong Lyman-$\alpha$ line
emission.  
This theoretical prediction was only confirmed thirty year later,  not
on first generation galaxies but on distant relatively young systems. 
Currently, Lyman Alpha Emitting (LAE) galaxies are commonly targetted
in surveys. 
The presence of the Ly-$\alpha$ emission line provides confirmation of
the distance of a galaxy and also provides clues about the stellar
population powering the emission.

Ly-$\alpha$ emission is not exclusive of distant galaxies. 
Any galaxy with low dust content and ongoing star formation has the
right conditions to show this line.  
There are, for instance,  local Universe surveys that target
Ly-$\alpha$ emission in nearby dwarf star forming galaxies. 
\cite{LARS}. 
A nearby LAE sample allow for the study of other indicators that
might be more difficult to obtain for distant galaxies such as
morphology, dust attenuation, neutral hydrogen contents and ionization
state.

However, the physical interpratation of Ly-$\alpha$ observations is
not straightforward. 
This is due to the resonant nature of the Ly-$\alpha$ line. 
A Ly-$\alpha$ photon followw a difussion-like process before escaping
the galaxy or being absorbed by dust. 
The reulting line profile becomes sensitive to the dynamical, chemical
and thermal conditions in the interstellar medium. 
Only in very few cases of high symmetry there are analytical solutions
available. 
Interpretation of observations require state-of-the-art models use
montecarlo radiative transfer simulations.

Recent computational models \cite{GaravitoCamargo2014} have
shown that rotation also imprints an effect on the Lyman-alpha
morphology.  
The most important consequence of rotation is that spherical 
symmetry is broken.
The line morphology now depends on the viewing angle respect to the
rotation axis.  
For a line of sight perpendicular to the rotation axis the intensity
and the line center and the line width increase with rotational
velocity.  
When the rotational velocity is close to the half-line width of the
static line, the line becomes single peaked as it is observed in
\tol, a unique feature that other theoretical models find
impossible to reproduce.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{tolfit.png}
\caption{{\bf Symmetric triple peaked Ly-$\alpha$ emission of \tol.}
  Dots correspond to the observational data. The line shows the result
of our best model from a full radiative transfer simulation. The best
parameters correspond to a rotational velocity of $300$km s$^{-1}$,
neutral Hydrogen optical depth of $10^{7}$ and a viewing angle of
$65^{\circ}$ measured from the rotation axis.} 
\end{center}
\end{figure}

>>>>

\begin{document}
\pagestyle{empty}
\noindent
\textbf{Þú ert jörðin}
\\
\\
JEFR$^{1}$, MCRG$^1$, JNGC$^2$, MD$^3$
\\
\\
\scriptsize
{$^1$ Bogota
\\
$^2$ Tucson
\\
$^3$ Oslo
\normalsize
\\
\\
\textbf{
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
features in its Lyman-$\alpha$ emission thad had evaded theoretical
interpretation so far. 
We show that these special features, a symmetric triple peaked
emission line, can be explained by gas bulk rotation.  
We find that the Lyman-$\alpha$ emission region in \tol\ should
have a rotational velocity of $V_{r}=300$ km s$^{-1}$ and a neutral
Hydrogen column density of $\log N_{HI} / \mathrm{atoms\ cm}^{-2} =
  20.5$.   
Using archival observational information about \tol\ we find that the
diameter for that emission region diameter should be in the range of
$110$ pc $<D<$ $340$ pc and its total dynamical mass   should be
between $2.1\times 10^{9}$M$_{\odot}$ $<M_D<$  $6.6\times
10^{9}$M$_{\odot}$.  
This dynamical mass is at least $16\pm 9$ times
larger than the neutral mass hydrogen.
We argue that a possibility to explain the excess in
dynamical mass is the presence of a super-massive black hole. }

The first generation of galaxies trace our cosmic origins. 
They were the first steps in the evolution of galaxies such as the Milky
Way. 
In the standard Big Bang cosmology the only elements that were
created in the nucleosynthesis process were Hydrogen, Helium and
Lithium. 
Heavier elemets must have been created in stellar evolution process. 
Therefore, we expect the first generation of
galaxies to be metal free and rich in Hydrogen. 
This kind of primordial galaxies have not been detected yet. 
However, dwarf star forming galaxies with a low metallicity content
are seen as templates to understand the early galaxy evolution process.

Almost fifty years ago \cite{PartridgePeebles} it was realized that
young galaxies could be detected through a strong Lyman-$\alpha$ line
emission.  
This theoretical prediction was only confirmed thirty year later on
distant relatively young, not primordial, galaxies.
Currently Lyman Alpha Emitting (LAE) galaxies are commonly targetted
in surveys. 
The presence of the Ly-$\alpha$ emission line provides confirmation of
the distance of a galaxy while provides clues about the stellar
population and inter-stellar medium conditions regulating the
Ly-$\alpha$ emission.

The Ly-$\alpha$ emission line is not exclusive of distant galaxies. 
Any galaxy with low dust content and ongoing star formation has the
right conditions to show this line.  
There are, for instance,  local Universe surveys that target
Ly-$\alpha$ emission in nearby dwarf star forming galaxies. 
\cite{LARS}. 
The study of nearby LAE samples has allowd the study of other
indicators that might be more difficult to obtain for distant galaxies
such as morphology, dust attenuation, neutral hydrogen contents and
ionization state.

However, the physical interpratation of Ly-$\alpha$ observations is
not straightforward. 
This is due to the resonant nature of the Ly-$\alpha$ line. 
A Ly-$\alpha$ photon follows a difussion-like process before escaping
the galaxy or being absorbed by dust. 
The resulting line profile becomes sensitive to the dynamical, chemical
and thermal conditions in the interstellar medium. 
There are very few analytical tools available to interpret the
Ly-$\alpha$ line.
They are applicable only in very few cases of highly symmetrical
conditions, which are hardly met in real astrophysical systems.

For these reasons the interpretation of Ly-$\alpha$ observations
require state-of-the-art montecarlo radiative transfer simulations.   
Recent advances in these computational models
\cite{GaravitoCamargo2014} have shown that galaxy rotation imprints an
effect on the Lyman-$\alpha$ line.
The most important consequence of rotation is that even for a
spherical gas distribution, the line morphology now depends on the
viewing angle respect to the rotation axis.  
For a line of sight perpendicular to the rotation axis the intensity
and the line center and the line width increase with rotational
velocity. 
When the rotational velocity is close to the half-line width of the
static line the line becomes single peaked, as it is observed in
\tol, a unique feature that other theoretical models find
difficult to reproduce without invoking complex gas distributions.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{tolfit.pdf}
\caption{{\bf Symmetric triple peaked Ly-$\alpha$ emission of \tol.}
  Dots correspond to the observational data. The line shows the result
of our best model from a full radiative transfer simulation. The best
parameters correspond to a rotational velocity of $300$km s$^{-1}$,
neutral Hydrogen optical depth of $10^{7}$ and a viewing angle of
$65^{\circ}$ measured from the rotation axis.} 
\end{center}
\end{figure}
